---
title: "metrics 2 paper"
author: "Murphy Li"
date: "2023-12-14"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
```{r}
library(pacman)
p_load(tidyverse, lfe, stargazer, lubridate, lmtest)

```

load data
```{r}
df<-read.csv('data.csv')%>%
  select(-c('Series.Code'))%>%
  rename(variable=Series.Name,
         country=Country.Name,
         code=Country.Code)



names(df)<-gsub("\\..*","",names(df))
names(df)<-gsub("X","",names(df))
```

pivot longer to change years to rows
```{r}
df<-df%>%
  pivot_longer(!c('variable', 'country', 'code'), names_to='year', values_to='value')



#make year and values numeric
df$year<-as.numeric(df$year)
df$value<-as.numeric(df$value)
#df$value[df$value=='']<-NA




```

pivot wider to change variable column to columns
```{r}
df<-df%>%
  distinct%>%
  pivot_wider(names_from=variable, values_from=value)
```

simplify column names
```{r}

names(df)[4]<-'gdpgr'
names(df)[5]<-'gdppercap'
names(df)[6]<-'urate'
names(df)[7]<-'exrate'
names(df)[8]<-'tarrate'
names(df)[9]<-'hightechexpr'
names(df)[10]<-'healthexp'
names(df)[11]<-'pop'
names(df)[12]<-'labpartr'

df <- df %>% mutate(pop_sq = pop^2)
```

 [1] "country"                                                                                        
 [2] "code"                                                                                           
 [3] "year"                                                                                           
 [4] "GDP growth (annual %)"                                                                          
 [5] "GDP per capita (current US$)"                                                                   
 [6] "Unemployment, total (% of total labor force) (national estimate)"                               
 [7] "Real effective exchange rate index (2010 = 100)"                                                
 [8] "Tariff rate, applied, simple mean, all products (%)"                                            
 [9] "High-technology exports (% of manufactured exports)"                                            
[10] "Current health expenditure (% of GDP)"                                                          
[11] "Population ages 15-64 (% of total population)"                                                  
[12] "Labor force participation rate, total (% of total population ages 15-64) (modeled ILO estimate)"


dummy for covid
```{r}
df$covid<-0
df$covid[df$year==2020]<-1
```




make factor variable for identifying low, low mid, up mid, high income countries 
```{r}
df$income <- ifelse(df$gdppercap > 12696, "High",
                         ifelse(df$gdppercap > 4096, "Upper mid",
                                ifelse(df$gdppercap > 1045, "Lower mid", "Low")))
df$income<-as.factor(df$income)


```
Low-income countries: GNI per capita of $1,045 or less.
Lower-middle-income countries: GNI per capita between $1,046 and $4,095.
Upper-middle-income countries: GNI per capita between $4,096 and $12,695.
High-income countries: GNI per capita of $12,696 or more.

summary statistics
```{r}
#mean table
mean_sum_stat <- df %>%
  group_by(income) %>%
  summarize(
    gdp_gr = round(mean(gdpgr, na.rm = TRUE), 1),
    gdp_per_cap = round(mean(gdppercap, na.rm = TRUE), 1),
    urate = round(mean(urate, na.rm = TRUE), 1),
    exrate = round(mean(exrate, na.rm = TRUE), 1),
    tar_rate = round(mean(tarrate, na.rm = TRUE), 1),
    HT_export = round(mean(hightechexpr, na.rm = TRUE), 1),
    health_exp = round(mean(healthexp, na.rm = TRUE), 1),
    pop = round(mean(pop, na.rm = TRUE), 1),
    labpart_r = round(mean(labpartr, na.rm = TRUE), 1)
  )
mean_sum_stat$income<-as.character(mean_sum_stat$income)

mean_sum_stat<-mean_sum_stat[-5,]

stargazer(mean_sum_stat, summary=F)
```


eda and graphs
```{r}
#all country


df%>%
  ggplot(aes(x=year, y=gdpgr, color=country))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Real GDP Growth Rate of all 217 countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#average by income level
df%>%
  select(c('year', 'gdpgr', 'income'))%>%
  na.omit()%>%
  group_by(year, income)%>%
  summarize(gdpgr=mean(gdpgr))%>%
  ggplot(aes(x=year, y=gdpgr, color=income))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  labs(title='Average Real GDP Growth Rate of Countries by Income level')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#only high
df%>%
  select(c('year', 'gdpgr', 'income'))%>%
  filter(income=='High')%>%
  na.omit()%>%
  group_by(year, income)%>%
  summarize(gdpgr=mean(gdpgr))%>%
  ggplot(aes(x=year, y=gdpgr, color=income))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Average Real GDP Growth Rate of High Income Countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#only low
df%>%
  select(c('year', 'gdpgr', 'income'))%>%
  filter(income=='Low')%>%
  na.omit()%>%
  group_by(year, income)%>%
  summarize(gdpgr=mean(gdpgr))%>%
  ggplot(aes(x=year, y=gdpgr, color=income))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Average Real GDP Growth Rate of Low Income Countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#only lower mid
df%>%
  select(c('year', 'gdpgr', 'income'))%>%
  filter(income=='Lower mid')%>%
  na.omit()%>%
  group_by(year, income)%>%
  summarize(gdpgr=mean(gdpgr))%>%
  ggplot(aes(x=year, y=gdpgr, color=income))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Average Real GDP Growth Rate of Lower Middle Income Countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#only upper mid
df%>%
  select(c('year', 'gdpgr', 'income'))%>%
  filter(income=='Upper mid')%>%
  na.omit()%>%
  group_by(year, income)%>%
  summarize(gdpgr=mean(gdpgr))%>%
  ggplot(aes(x=year, y=gdpgr, color=income))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Average Real GDP Growth Rate of Upper Middle Income Countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#all countries in upper mid
df%>%
  select(c('year', 'gdpgr', 'income', 'country'))%>%
  filter(income=='Upper mid')%>%
  na.omit()%>%
  ggplot(aes(x=year, y=gdpgr, color=country))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Real GDP Growth Rate of Upper Middle Income Countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#all countries in lower mid
df%>%
  select(c('year', 'gdpgr', 'income', 'country'))%>%
  filter(income=='Lower mid')%>%
  na.omit()%>%
  ggplot(aes(x=year, y=gdpgr, color=country))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Real GDP Growth Rate of Lower Middle Income Countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#all countries in high
df%>%
  select(c('year', 'gdpgr', 'income', 'country'))%>%
  filter(income=='High')%>%
  na.omit()%>%
  ggplot(aes(x=year, y=gdpgr, color=country))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Real GDP Growth Rate of High Income Countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')

#all countries in low
  select(c('year', 'gdpgr', 'income', 'country'))%>%
  filter(income=='Low')%>%
  na.omit()%>%
  ggplot(aes(x=year, y=gdpgr, color=country))+
  geom_line()+
  theme_bw()+
  geom_vline(xintercept=2020)+
  theme(legend.position = "none")+
  labs(title='Real GDP Growth Rate of Low Income Countries')+
  ylab('Real GDP Growth Rate (%)')+
  xlab('Year')
```

model 1 with all variables
```{r}
#looking at all countries
mod1<-felm(gdpgr~urate + exrate + tarrate + hightechexpr + healthexp + pop+ pop_sq+ labpartr +  covid|country |0|country, df )
mod2<-felm(gdpgr~urate + tarrate + hightechexpr + healthexp +  labpartr + covid|country|0|country, df )
mod3<-felm(gdpgr~urate + exrate +  hightechexpr + healthexp + pop +  covid |country |0|country, df )
mod4<-felm(gdpgr~urate  +  hightechexpr + healthexp + pop +  covid |country |0|country, df )

#incorporating their wealth (looking at gdp per cap)
mod5<-felm(gdpgr~urate + exrate + tarrate + hightechexpr + healthexp + pop+ pop_sq+ labpartr +  covid *income|country |0|country, df )
mod6<-felm(gdpgr~urate + tarrate + hightechexpr + healthexp +  labpartr + covid*income|country|0|country, df )
mod7<-felm(gdpgr~urate + exrate +  hightechexpr + healthexp + pop +  covid *income|country |0|country, df )
mod8<-felm(gdpgr~urate  +  hightechexpr + healthexp + pop +  covid * income|country|0|country, df )


summary(mod1, robust=T)
AIC(mod1)
summary(mod2, robust=T)
AIC(mod2)
summary(mod3, robust=T)
AIC(mod3)
summary(mod4, robust=T)
AIC(mod4)
summary(mod5, robust=T)
AIC(mod5)
summary(mod6, robust=T)
AIC(mod6)
summary(mod7, robust=T)
AIC(mod7)
summary(mod8, robust=T)
AIC(mod8)

#unrestricted model
stargazer(mod1, mod5)

#restricted model with no interactive term
stargazer(mod2, mod3, mod4)

#restricted model with interactive term
stargazer(mod6, mod7, mod8)

summary(mod1, robust=T)
AIC(mod1)
summary(mod5, robust=T)
AIC(mod5)





```
model 6 is best


Eigen values for cor matrix for models to check for multicolinearity (small is yes multicolinearity)
```{r}
eigen(cor(model.matrix(mod1)))$values
eigen(cor(model.matrix(mod2)))$values
eigen(cor(model.matrix(mod3)))$values
eigen(cor(model.matrix(mod4)))$values
eigen(cor(model.matrix(mod5)))$values
eigen(cor(model.matrix(mod6)))$values
eigen(cor(model.matrix(mod7)))$values
eigen(cor(model.matrix(mod8)))$values
```
models 1, 5, 7?,8  may suffer from multicolinearity, cannot run VIF test since model has no intercept

residual plots for the 4 models
```{r}
#MOD 1
# Extract residuals and fitted values
residuals <- resid(mod1)
fitted_values <- fitted(mod1)

# Create a data frame for ggplot
plot_data <- data.frame(Fitted = fitted_values, Residuals = residuals)

# Create a residual plot
ggplot(plot_data, aes(x = gdpgr, y = gdpgr.1)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Base model with no interactive term", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

#MOD 2
# Extract residuals and fitted values
residuals <- resid(mod2)
fitted_values <- fitted(mod2)

# Create a data frame for ggplot
plot_data <- data.frame(Fitted = fitted_values, Residuals = residuals)

# Create a residual plot
ggplot(plot_data, aes(x = gdpgr, y = gdpgr.1)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual restricted model 1 with no interactive term", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

#MOD 3
# Extract residuals and fitted values
residuals <- resid(mod3)
fitted_values <- fitted(mod3)

# Create a data frame for ggplot
plot_data <- data.frame(Fitted = fitted_values, Residuals = residuals)

# Create a residual plot
ggplot(plot_data, aes(x = gdpgr, y = gdpgr.1)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Restricted model 2 with no interactive term", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

#MOD 4
# Extract residuals and fitted values
residuals <- resid(mod4)
fitted_values <- fitted(mod4)

# Create a data frame for ggplot
plot_data <- data.frame(Fitted = fitted_values, Residuals = residuals)

# Create a residual plot
ggplot(plot_data, aes(x = gdpgr, y = gdpgr.1)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Restricted model 3 with no interactive term", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

#MOD 5
# Extract residuals and fitted values
residuals <- resid(mod5)
fitted_values <- fitted(mod5)

# Create a data frame for ggplot
plot_data <- data.frame(Fitted = fitted_values, Residuals = residuals)

# Create a residual plot
ggplot(plot_data, aes(x = gdpgr, y = gdpgr.1)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Base model with interactive term", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

#MOD 6
# Extract residuals and fitted values
residuals <- resid(mod6)
fitted_values <- fitted(mod6)

# Create a data frame for ggplot
plot_data <- data.frame(Fitted = fitted_values, Residuals = residuals)

# Create a residual plot
ggplot(plot_data, aes(x = gdpgr, y = gdpgr.1)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Restricted model 1 with interactive term", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

#MOD 7
# Extract residuals and fitted values
residuals <- resid(mod7)
fitted_values <- fitted(mod7)

# Create a data frame for ggplot
plot_data <- data.frame(Fitted = fitted_values, Residuals = residuals)

# Create a residual plot
ggplot(plot_data, aes(x = gdpgr, y = gdpgr.1)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Restricted model 2 with interactive term", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

#MOD 8
# Extract residuals and fitted values
residuals <- resid(mod8)
fitted_values <- fitted(mod8)

# Create a data frame for ggplot
plot_data <- data.frame(Fitted = fitted_values, Residuals = residuals)

# Create a residual plot
ggplot(plot_data, aes(x = gdpgr, y = gdpgr.1)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Restricted model 3 with interactive term", x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```
# slight observable heteroskedasticity, not apparent. Use white test to check

White and bp test for heteroskedasticity
```{r}
bptest(mod1)
bptest(mod2)
bptest(mod3)
bptest(mod4)
bptest(mod5)
bptest(mod6)
bptest(mod7)
bptest(mod8)


```
BP test suggest all models are heteroskedastic except for model 6

plot residuals to check distribution
```{r}
# mod 1
residuals_mod1 <- residuals(mod1)

ggplot(data.frame(residuals = residuals_mod1), aes(x = gdpgr)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "white", fill = "lightblue", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(title = "Histogram with Bell Curve for Residuals Mod 1") +
  theme_minimal()

# mod 2
residuals_mod2 <- residuals(mod2)

ggplot(data.frame(residuals = residuals_mod2), aes(x = gdpgr)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "white", fill = "lightblue", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(title = "Histogram with Bell Curve for Residuals Mod 2") +
  theme_minimal()

# mod 3
residuals_mod3 <- residuals(mod3)

ggplot(data.frame(residuals = residuals_mod3), aes(x = gdpgr)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "white", fill = "lightblue", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(title = "Histogram with Bell Curve for Residuals Mod 3") +
  theme_minimal()

# mod 4
residuals_mod4 <- residuals(mod4)

ggplot(data.frame(residuals = residuals_mod4), aes(x = gdpgr)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "white", fill = "lightblue", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(title = "Histogram with Bell Curve for Residuals Mod 4") +
  theme_minimal()

# mod 5
residuals_mod5 <- residuals(mod5)

ggplot(data.frame(residuals = residuals_mod5), aes(x = gdpgr)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "white", fill = "lightblue", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(title = "Histogram with Bell Curve for Residuals Mod 1") +
  theme_minimal()

# mod 6
residuals_mod6 <- residuals(mod6)

ggplot(data.frame(residuals = residuals_mod6), aes(x = gdpgr)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "white", fill = "lightblue", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(title = "Histogram with Bell Curve for Residuals Mod 2") +
  theme_minimal()

# mod 3
residuals_mod7 <- residuals(mod7)

ggplot(data.frame(residuals = residuals_mod7), aes(x = gdpgr)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "white", fill = "lightblue", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(title = "Histogram with Bell Curve for Residuals Mod 3") +
  theme_minimal()

# mod 4
residuals_mod8 <- residuals(mod8)

ggplot(data.frame(residuals = residuals_mod8), aes(x = gdpgr)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "white", fill = "lightblue", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(title = "Histogram with Bell Curve for Residuals Mod 4") +
  theme_minimal()
```

Normality test
```{r}
shapiro.test(mod1$residuals)
shapiro.test(mod2$residuals)
shapiro.test(mod3$residuals)
shapiro.test(mod4$residuals)
shapiro.test(mod5$residuals)
shapiro.test(mod6$residuals)
shapiro.test(mod7$residuals)
shapiro.test(mod8$residuals)
```
residuals are not normal, there could be omitted variable bias
